---
title: "QC Report"
date: "`r format(Sys.time(), '%d %B, %Y')`"
author: "`r Sys.info()[[7]]`"
runtime: shiny
output:
  html_document:
    toc: true
    toc_float: true
    toc_collapsed: true
    theme: yeti
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, fig.align="center", warning=FALSE, message=FALSE)
```

```{r libraries, include=FALSE}
library("ggplot2")
library("tidyverse")
library("optparse")
library("ggpubr")
library("knitr")
library("fastqcr")
library("shiny")
library(DT)
```

Report generated based on data located in `r getwd()` \  

```{r samtool, include=FALSE}
## Samtools ##

sam_tools <- read_delim(file = "multiqc_data/multiqc_samtools_stats.txt", delim = '\t', col_names = TRUE)
sam_tools <- dplyr::select(sam_tools, c("Sample", "raw_total_sequences", "reads_mapped", "reads_unmapped"))
sam_tools <- gather(sam_tools, "Mapped/Unmapped", "Number", 3:4)
sam_tools$`Mapped/Unmapped` <- ifelse(sam_tools$`Mapped/Unmapped` == "reads_mapped", "Mapped", "Unmapped" )

sam_tools$Percentages <- 100*(sam_tools$Number / sam_tools$raw_total_sequences)
sam_tools$Sample <- gsub("\\.stats$", "", sam_tools$Sample )
```

Samples: `r unique(sam_tools$Sample)` \n  

Report generated at: `r format(Sys.time(), "%X %x ")` \n 

# General stats
MultiQC general stats from Bowtie, featureCounts, FastQC and Samtools. 
```{r, general_stats, echo = FALSE}
# General stats
general <- read_delim(file = "multiqc_data/multiqc_general_stats.txt", delim = '\t', col_names = TRUE)
colnames(general) <- c("Sample Name", "Bowtie: % Reads Aligned", "Bowtie: Reads Aligned", "FC: % Assigned", "FC: Assigned", "FastQC: % Duplicates", "FastQC: % GC", "FastQC: Average Sequence Length", "FastQC: % Fails", "FastQC: Total Sequences", "Samtools: % Error Rate", "Samtools: Non Primary Allignments", "Samtools: Reads Mapped", "Samtools: % Reads Mapped", "Samtools: Raw Total Sequences")
general$`Samtools: % Error Rate` <- as.numeric(general$`Samtools: % Error Rate`)*100
#general <- general %>% replace(is.na(.), "-")
datatable(general) 
```
FC: featureCounts

# Samtools 

SAM (Sequence Alignment/Map) format is a generic format for storing large nucleotide sequence alignments.

SAM Tools provide various utilities for manipulating alignments in the SAM format, including sorting, merging, indexing and generating alignments in a per-position format (hosted on github: https://github.com/samtools/samtools)

## Reads mapped stats {.tabset .tabset-pills}

### Number of reads
Mapped vs unmapped reads
```{r samtools_plot1, include=TRUE, echo =FALSE}
# Plots
samtools_plot <- ggplot(data=sam_tools, aes(x=Sample, y=Number, fill=`Mapped/Unmapped`)) +
  geom_bar(stat="identity") + coord_flip() + fill_palette("lancet")

print(samtools_plot)
```

### Percentages
Mapped vs unmapped reads
```{r, samtools_plot2, include = TRUE, echo=FALSE}
samtools_plot_percent <- ggplot(data=sam_tools, aes(x=Sample, y=Percentages, fill=`Mapped/Unmapped`)) +
  geom_bar(stat="identity") + coord_flip() + fill_palette("lancet")

print(samtools_plot_percent)
```

## Alignment Metrics 
Alignment metrics from samtools stats

```{r, kable_alignment_metrics, echo = FALSE, fig.align='center'}
sam_table <- read_tsv("multiqc_data/multiqc_samtools_stats.txt")
sam_table$Sample <- gsub("\\.stats$", "", sam_table$Sample)
sam_table_subset <- sam_table %>% dplyr::select(contains("Sample"),  contains("total_sequences"), contains("reads"), contains("bases"), contains("pairs"), -contains("percent")) %>%
  rename(total_sequences = raw_total_sequences, `mapped_paired` = reads_mapped_and_paired, `MQ0_reads` = reads_MQ0 ) %>% rename_all(~sub('reads_', '', .x)) %>%
  rename_all(~sub('^pairs[^_]*_[^_]*_', '', .x))
datatable(sam_table_subset)
```

## Mapped reads per contig {.tabset .tabset-pills}
The samtools idxstats tool counts the number of mapped reads per chromosome / contig. Chromosomes with < 0.1% of the total aligned reads are omitted from this plot

```{r mapped_reads_contig, include = FALSE}
contig <- read_delim(file = "multiqc_data/multiqc_samtools_idxstats.txt", delim = '\t', col_names = TRUE)
contig <- contig %>% replace(is.na(.), 0)
contig$total <- rowSums(contig[, -c(1)])

# Chromosomes with < 0.1% of the total aligned reads are omitted from this plot
contig <- contig %>% select_if(function(x) is.character(x) || sum(x) > sum(contig$total)*0.001)

contig <- gather(contig, "Chromosome Name", "Count", -c(1, ncol(contig)))

contig$Percentages <- 100*(contig$Count / contig$total)

# Normalised
contig_plot_norm <- ggplot(contig, aes(x = `Chromosome Name`, y = Percentages, group=Sample)) + geom_line(aes(color=Sample))  +
  theme_bw() + theme(axis.text.x = element_text(angle = 45, hjust = 1)) 
contig_plot_norm <- set_palette(contig_plot_norm, "lancet")

# Count
contig_plot_count <- ggplot(contig, aes(x = `Chromosome Name`, y = Count, group=Sample)) + geom_line(aes(color=Sample)) + 
  theme_bw() + theme(axis.text.x = element_text(angle = 45, hjust = 1)) 
contig_plot_count <- set_palette(contig_plot_count, "lancet")
```

### Normalised counts
```{r mapped_reads_contig_norm, echo = FALSE, fig.height=5, fig.width=10}
print(contig_plot_norm)
```

### Counts
```{r mapped_reads_count, echo = FALSE, fig.height=5, fig.width=10}
print(contig_plot_count)
```


# Bowtie {.tabset .tabset-pills}

```{r bowtie_intro, include = FALSE}
bowtie <- read_delim(file = "multiqc_data/multiqc_bowtie1.txt", delim = '\t', col_names = TRUE)
bowtie <- dplyr::select(bowtie, -contains("percentage"))
bowtie <- gather(bowtie, "Aligned/Not aligned", "Number", 3:4)
bowtie$`Aligned/Not aligned` <- ifelse(bowtie$`Aligned/Not aligned` == "reads_aligned", "Aligned", "Not Aligned")
bowtie$Percentages <- 100*(bowtie$Number / bowtie$reads_processed)

# Number Plot
bowtie_plot <- ggplot(data=bowtie, aes(x=Sample, y=Number, fill=`Aligned/Not aligned`)) +
  geom_bar(stat="identity") + coord_flip() + fill_palette("lancet")

# Percentage plot
bowtie_plot_percent <- ggplot(data=bowtie, aes(x=Sample, y=Percentages, fill=`Aligned/Not aligned`)) +
  geom_bar(stat="identity") + coord_flip() + fill_palette("lancet")
```

## Number of Reads
```{r number_bowtie_plot, echo = FALSE}
print(bowtie_plot)
```

## Percentages
```{r percent_bowtie_plot, echo = FALSE}
print(bowtie_plot_percent)
```

# Feature Counts 

Subread featureCounts is a highly efficient general-purpose read summarization program that counts mapped reads for genomic features. Feature counts obtains data from bam files located in mapping.dir.

## Gene IDs {.tabset .tabset-pills}

```{r feature_counts, include = FALSE}
file_paths = list.files(pattern='_small.tsv$', recursive=TRUE, full.names = TRUE)
num_samples <- length(file_paths)
feature_count_tibble_all_samples = NULL

for(i in 1:num_samples){
  path <- file_paths[i]
  feature_count_tibble <-read_tsv(path, col_names = FALSE, skip = 2)
  colnames(feature_count_tibble) <- c("GeneID", "NumberMapped")
  # Get sample name 
  sample <- stringr::str_match(path, ".*/.*/(.*?)/")[2]
  feature_count_tibble$Sample <- sample
  total_mapped <- sum(feature_count_tibble$NumberMapped)
  feature_count_tibble$Percentages <- 100*(feature_count_tibble$NumberMapped / total_mapped)
  feature_count_tibble_all_samples = bind_rows(feature_count_tibble_all_samples,feature_count_tibble)
}
fc_tibble_all_samples_nonzero <- dplyr::filter(feature_count_tibble_all_samples, NumberMapped != 0)
fc_tibble_all_samples_nonzero<- fc_tibble_all_samples_nonzero[,c(3,1,2,4)]

feature_count_plot <- ggplot(data=fc_tibble_all_samples_nonzero, aes(x=Sample, y=NumberMapped, fill=`GeneID`)) +
  geom_bar(stat="identity") + coord_flip() + fill_palette("lancet")

feature_count_plot_percent <- ggplot(data=fc_tibble_all_samples_nonzero, aes(x=Sample, y=Percentages, fill=`GeneID`)) +
  geom_bar(stat="identity") + coord_flip() + fill_palette("lancet")
  
```

### Number
```{r fc_plot, echo =FALSE, fig.align='center'}
print(feature_count_plot)
```

### Percent
```{r fc_plot_percent, echo =FALSE, fig.align='center'}
print(feature_count_plot_percent)
```

## Feature Count Table
```{r, fc_table, echo=FALSE, fig.align='center'}
datatable(fc_tibble_all_samples_nonzero) 
```

## Assigned/ Unassigned {.tabset .tabset-pills}

```{r fc_mapped, include = FALSE}
fc_tibble <- read_tsv("multiqc_data/multiqc_featureCounts.txt", col_names = TRUE, skip = 0)
fc_tibble <- fc_tibble[, !names(fc_tibble) %in% c("percent_assigned")]
fc_tibble <- gather(fc_tibble, "Status", "Number", -c(1,2) )
fc_tibble$Percentages <- 100*(fc_tibble$Number/fc_tibble$Total)
fc_tibble <- fc_tibble[fc_tibble$Number != 0,]

feature_count_plot <- ggplot(data=fc_tibble, aes(x=Sample, y=Number, fill=`Status`)) +
  geom_bar(stat="identity") + coord_flip() + fill_palette("lancet")

feature_count_plot_percent <- ggplot(data=fc_tibble, aes(x=Sample, y=Percentages, fill=`Status`)) +
  geom_bar(stat="identity") + coord_flip() + fill_palette("lancet")
  
fc_tibble_final <- fc_tibble[, !names(fc_tibble) %in% c("Total")]
```


### Number of reads
```{r fc_plot2, echo = FALSE}
print(feature_count_plot)
```
`r datatable(fc_tibble_final)`

### Percentages
```{r fc_plot3, echo = FALSE}
print(feature_count_plot_percent)
```

`r datatable(fc_tibble_final)`


# Fast QC 
FastQC is used to perfrom quality control checks on raw high throughput sequencing data. 

``` {r, fast_QC, include = FALSE}

## Fill data frames using fastqcr module and rbind samples together
mean_quality_scores <- NULL
per_sequence_qs <- NULL
per_sequence_gc <- NULL
per_base_n_content <- NULL
sequence_duplication_levels <- NULL
sequence_duplication_levels <- NULL
overrepresented_sequences <- NULL
adapter_content <- NULL
file_paths = list.files(pattern='fastqc.zip$', recursive=TRUE, full.names = TRUE)
for(i in 1:length(file_paths)){
  sample_name <- sub(".*fastqc_pre.dir/(.*)?_fastqc\\.zip", "\\1", file_paths[i])
  qc_sample <-qc_read(file= file_paths[i], modules = "all")
  # K mer content and total deduplicated percentage removed and sample name added
  qc_sample <- Map(cbind, qc_sample[1:12], Sample = sample_name)

  mean_quality_scores <- rbind(mean_quality_scores, qc_sample[[3]])
  per_sequence_qs <- rbind(per_sequence_qs, qc_sample[[5]] )
  per_sequence_gc <- rbind(per_sequence_gc,qc_sample[[7]])
  per_base_n_content <- rbind(per_base_n_content, qc_sample[[8]])
  sequence_duplication_levels <- rbind(sequence_duplication_levels, qc_sample[[10]])
  adapter_content <- rbind(adapter_content, qc_sample[[12]])
  
  overrepresented_sequences_subset <- tibble(group = c("Top over-represented sequence", "Sum of remaining over-represented sequences"), Percentages = c(qc_sample[[11]]$Percentage[1], sum(qc_sample[[11]]$Percentage) - qc_sample[[11]]$Percentage[1]), Sample = sample_name )
  overrepresented_sequences <- rbind(overrepresented_sequences, overrepresented_sequences_subset)
}
```

```{r, initialise_FQ_plots, include = FALSE}
## Initialise plots (to be printed individually)
mean_q_scores_plot <- ggplot(data=mean_quality_scores, aes(x=Base, y=Mean, color=Sample)) +
  geom_line(stat="identity") + labs(x = "Position (bp)", y= "Mean phred score") + expand_limits(y=0) +
  theme_bw() 
mean_q_scores_plot <- set_palette(mean_q_scores_plot, "lancet")

per_sequence_qs_plot <- ggplot(data=per_sequence_qs, aes(x=Quality, y=Count, color=Sample)) +
  geom_line(stat="identity") + labs(x = "Mean Sequence Quality (Phred Score)", y= "Count") + expand_limits(y=0) + theme_bw() 
per_sequence_qs_plot <- set_palette(per_sequence_qs_plot, "lancet")

per_sequence_gc_plot <- ggplot(data=per_sequence_gc, aes(x=`GC Content`, y=Count, color=Sample)) +
  geom_line(stat="identity") + labs(x = "% GC Content", y= "Count") + theme_bw()   
per_sequence_gc_plot <- set_palette(per_sequence_gc_plot, "lancet")

per_base_n_content_plot <- ggplot(data=per_base_n_content, aes(x=Base, y=`N-Count`, color=Sample)) +
  geom_line(stat="identity") + labs(x = "Position in read (bp)", y= "Percentage N-Count") + theme_bw()
per_base_n_content_plot <- set_palette(per_base_n_content_plot, "lancet")


overrepresented_sequences_plot <- ggplot(data=overrepresented_sequences, aes(x=Sample, y=Percentages, fill=group)) +
  geom_bar(stat="identity") + labs(y = "Percentage of total sequences") + coord_flip() + fill_palette("lancet") + theme_bw() + ylim(0,100) + theme(legend.position="bottom") + labs(fill = "")

## 
sequence_duplication_levels$`Duplication Level` <- factor(sequence_duplication_levels$`Duplication Level`, as.character(sequence_duplication_levels$`Duplication Level`[1:16]))

sequence_duplication_levels_plot <- ggplot(data=sequence_duplication_levels, aes(x=`Duplication Level`, y=`Percentage of total`, color = Sample, group =1)) + geom_line(stat="identity") + labs(x = "Sequence Duplication Level", y= "Percentage of Library") +ylim(0,100) +theme_bw()
sequence_duplication_levels_plot <- set_palette(sequence_duplication_levels_plot, "lancet") 

# Adapter content
# Cumulative
adapter_content_cumulative <- adapter_content
adapter_content_cumulative$total_adapter <- rowSums(adapter_content_cumulative[,2:6] )
adapter_content_cumulative <- adapter_content_cumulative %>% select(-contains(" Adapter"), -contains("Nextera"))
adapter_content_cumulative_plot <- ggplot(data= adapter_content_cumulative, aes(x= Position, y= total_adapter, color = Sample)) + geom_line() + theme_bw() + labs(x = "Position (bp)", y = "Percentage of Sequences")
adapter_content_cumulative_plot <- set_palette(adapter_content_cumulative_plot, "lancet")

# Each adapter type individually
adapter_content <- gather(adapter_content, "Adapter", "Percentage_of_sequences", 2:6)
adapter_content_plot <- ggplot(data= adapter_content, aes(x= Position, y= Percentage_of_sequences, color = Sample, fill = Sample, shape = Adapter, 
  group=interaction(Sample, Adapter))) + 
  geom_point() + geom_line() + theme_bw() + labs(x = "Position (bp)", y = "Percentage of Sequences")
adapter_content_plot <- set_palette(adapter_content_plot, "lancet")
```

## Mean quality scores
The mean quality value across each base position in the read.
```{r, QC_mean_plot, echo = FALSE}
print(mean_q_scores_plot)
```

## Per sequence quality scores
The number of reads with average quality scores. Shows if a subset of reads has poor quality
```{r, per_sequence_qs_plot, echo=FALSE}
print(per_sequence_qs_plot)
```

## Per sequence GC content
The average GC content of reads. Normal random library typically have a roughly normal distribution of GC content.
```{r, per_sequence_gc_plot, echo=FALSE}
print(per_sequence_gc_plot)
```

## Per base N content
If a sequencer is unable to make a base call with sufficient confidence then it will normally substitute an N rather than a conventional base. This plot shows the percentage of base calls at each position for which an N was called. The documentation states that you get a warning if the number of Ns is above 5%. 
```{r, per_base_n_count, echo=FALSE}
print(per_base_n_content_plot)
```

## Sequence Duplicaton Levels
The relative level of duplication found for every sequence.
```{r, sequence_duplication_levels_plot, echo=FALSE}
print(sequence_duplication_levels_plot)
```

## Overrepresented Sequences
The total amount of overrepresented sequences found in each library
```{r, overrepresented_sequences_plot, echo = FALSE}
print(overrepresented_sequences_plot)
```

## Adapter content {.tabset .tabset-pills}

### Specific adapter content 
The percentage count of the proportion of your library which has seen individual types of adaptor sequences at each position. 
```{r, adapter_content_plot, echo = FALSE}
print(adapter_content_plot)
```

### Adapter content cumulative
The cumulative percentage count of the proportion of your library which has seen each of the adapter sequences at each position. 
```{r, adapter_content_plot_cumsum, echo = FALSE}
print(adapter_content_cumulative_plot)
```

## Per base sequence content
### Shiny attempt
The proportion of each base position for which each of the four normal DNA bases has been called.
```{r, shiny, echo = FALSE}
file_paths = list.files(pattern='fastqc.zip$', recursive=TRUE, full.names = TRUE)
samples <- sub(".*fastqc_pre.dir/(.*)?_fastqc\\.zip", "\\1", file_paths)
shinyApp(
  ui = fluidPage(selectInput('my_dropdown','Choose Sample', samples),
                plotOutput('my_plot')),
  server = function(input, output) {
    
    op1 <- reactive({
      per_base_content <- qc_read(file= paste("fastqc_pre.dir/", input$my_dropdown,"_fastqc.zip", sep = ""), modules = "all")[[6]] %>% gather("Nucleotide Base", "Percentage", 2:5)
    })

    output$my_plot <- renderPlot({
          qc_sample <-qc_read(file= op1(), modules = "all")
    per_base_content <- qc_sample[[6]]
    per_base_content <- gather(per_base_content, "Nucleotide Base", "Percentage", 2:5)
      ggplot(data=per_base_content, aes(x=Base, y=Percentage, fill=`Nucleotide Base`)) +
        geom_bar(stat="identity") + labs(x = "Base Pair", title = input$my_dropdown)  + fill_palette("lancet")
    })
  }, 
  options = list(height = 500)
)

```